---
title: "metab"
format: html
editor: visual
---

## Positive mode

Load the required R packages and the positive ion mode metabolomics dataset.\

Load metabolomic data for positive ion mode. Set up consistent variable names to avoid overwriting issues later in the pipeline.

```{r}
# Load required packages
library(pmp)
library(SummarizedExperiment)
library(dplyr)
library(ggplot2)

# Load data
load("metabolomics_pos.RData")

# Re-assign to new variable names to avoid overwrite issues
features_info_pos <- features_info
samples_info_pos <- samples_info
colnames(features_info_pos)

```

Clean sample info and align metadata with the intensity matrix.

Build Summarized Experiment object for structured downstream analysis.

QC replicates "QC31", "QC32", "QC5_2" are in the data matrix but not in samples info (positive), therefore they were removed

```{r}
# Clean sample info to match data matrix
data_pos <- data_pos[, !(colnames(data_pos) %in% c("QC31", "QC32", "QC5_2"))]
samples_info_pos <- samples_info_pos[
  match(colnames(data_pos), samples_info_pos$id),
]
stopifnot(all(colnames(data_pos) == samples_info_pos$id))  # Must be TRUE

# Align rownames
rownames(features_info_pos) <- rownames(data_pos)
rownames(samples_info_pos) <- colnames(data_pos)

# Create SummarizedExperiment object
se_raw <- SummarizedExperiment(
  assays = list(counts = as.matrix(data_pos)),
  rowData = features_info_pos,
  colData = samples_info_pos
)




```

### Perform multi-step feature filtering:

Removing samples with too many missing values.

Keeping features detected in 90% of QC samples and all samples.

Removing unstable features with high RSD in QCs.

```{r}
# Filter: features in ≥90% of QC samples
se_filt_qc <- filter_peaks_by_fraction(
  df = se_raw,
  min_frac = 0.9,
  classes = se_raw$strain,
  method = "QC",
  qc_label = "QCs"
)


# Filter: remove features with RSD > 30% in QC
se_filt_rsd <- filter_peaks_by_rsd(
  df = se_filt_qc,
  max_rsd = 30,
  classes = se_filt_qc$strain,
  qc_label = "QCs"
)
```

PQN normalisation using QC samples as reference

Imputing missing values with KNN

Apply GLOG transformation to stabilize variance across intensities.

```{r}
# PQN Normalisation
se_pqn <- pqn_normalisation(
  df = se_filt_rsd,
  classes = se_filt_rsd$strain,
  qc_label = "QCs"
)

# Missing value imputation (KNN)
se_imputed <- mv_imputation(
  df = se_pqn,
  method = "knn"
)

# Save pre-GLOG version for comparison
se_pre_glog <- se_imputed  # 📌 This is our "before" snapshot

# GLOG Transformation
se_glog <- glog_transformation(
  df = se_imputed,
  classes = se_imputed$strain,
  qc_label = "QCs"
)

```

#### Visualising the transformations

Visualize optimal lambda for GLOG and histograms of intensity distributions before and after transformation.

Shows how the Sum of Squared Errors (SSE) changes with different lambda values used in the GLOG transformation, helping to identify the most variance-stabilizing parameter

```{r}

# Visualize lambda optimization
opt_lambda <- processing_history(se_glog)$glog_transformation$lambda_opt
glog_plot_optimised_lambda(
  df = se_glog,
  optimised_lambda = opt_lambda,
  classes = se_pre_glog$strain,
  qc_label = "QCs"
)

# Histograms: Before vs After GLOG
par(mfrow = c(1, 2))
hist(assay(se_pre_glog),
     breaks = 100, main = "Before GLOG", xlab = "Intensity", col = "tomato")
hist(assay(se_glog),
     breaks = 100, main = "After GLOG", xlab = "Transformed Intensity", col = "steelblue")

```

Transformation structure checks for further analysis

```{r}
# colnames(se_glog)
# rownames(se_glog)
# colnames(samples_info_pos)
```

### PCA

Load PCA libraries

```{r}
# Ensure necessary libraries are installed and loaded
if (!requireNamespace("pmp", quietly = TRUE)) install.packages("pmp")
if (!requireNamespace("ggplot2", quietly = TRUE)) install.packages("ggplot2")
if (!requireNamespace("ggfortify", quietly = TRUE)) install.packages("ggfortify")

```

##### Full Data PCA

Run PCA on normalised data to explore clustering, using metadata for visual grouping

```{r}

library(pmp)
library(ggplot2)
library(ggfortify)

# Assuming 'se_pos' is your SummarizedExperiment object after preprocessing
# Extract the assay data
data_matrix <- assay(se_glog)

# Perform PCA
pca_results <- prcomp(t(data_matrix), center = TRUE, scale. = TRUE)

# Create a data frame with PCA results and sample information
pca_data <- data.frame(
  PC1 = pca_results$x[, 1],
  PC2 = pca_results$x[, 2],
  Strain = se_glog$strain,
  Next2 = se_glog$Next2,
  Day = se_glog$day
)

ggplot(pca_data, aes(x = PC1, y = PC2, color = Strain, shape = Next2)) +
  geom_point(size = 3) +
  labs(
    title = "PCA of Metabolomics Data",
    x = "Principal Component 1",
    y = "Principal Component 2"
  ) +
  theme_minimal()

```

### Strain A next2 C vs B

Focusing on specific conditions

```{r}
# Load libraries (assumed already loaded earlier in pipeline)
library(FactoMineR)
library(factoextra)
library(ggplot2)

# Subset samples: Strain A with Next2 = B or C
idx_subset_pos <- which(se_glog$strain == "A" & se_glog$Next2 %in% c("B", "C"))
se_pos_subset <- se_glog[, idx_subset_pos]

# Prepare data and metadata
expr_pos_sub <- t(assay(se_pos_subset))                      # Transpose: samples = rows
meta_pos_sub <- as.data.frame(colData(se_pos_subset))       # Sample metadata

# Run PCA
pca_pos_res <- PCA(expr_pos_sub, graph = FALSE)
explained_var_pos <- round(pca_pos_res$eig[1:2, 2], 1)       # PC1 and PC2 variance %

# Format PCA coordinates with metadata
pca_coords_pos <- data.frame(pca_pos_res$ind$coord)
pca_coords_pos$Next2 <- meta_pos_sub$Next2
pca_coords_pos$Day <- meta_pos_sub$day


ggplot(pca_coords_pos, aes(x = Dim.1, y = Dim.2, color = Next2, shape = Day)) +
  geom_point(size = 3) +
  scale_color_manual(values = c("B" = "red", "C" = "blue")) +  # Flip the default colors
  labs(
    title = "PCA - Strain A (Next2 B vs C, Positive Mode)",
    x = paste0("PC1 (", explained_var_pos[1], "%)"),
    y = paste0("PC2 (", explained_var_pos[2], "%)")
  ) +
  theme_minimal()

```

### Likelihood ratio test (LRT)

Prepare data for LRT: Extract metadata and expression data, ensure variables are factors.

Run LRT per feature comparing full vs reduced models.

Adjust p-values (FDR) and identify top features.

```{r}
# Install necessary packages (if not already installed)
if (!requireNamespace("lmtest", quietly = TRUE)) install.packages("lmtest")

# Load libraries
library(lmtest)
library(SummarizedExperiment)
library(dplyr)

```

# Likelihood Ratio Testing

Preparing data and running for looped LRT

Question: Does adding Next2 to the model explain significantly more variation in intensity than day alone?

-   Nested model: `intensity ~ day`
-   Complex model: `intensity ~ Next2 + day`

### 1. **Subset the Data:**

```{r}
# Extract relevant samples: strain A with Next2 in B or C
idx <- which(colData(se_glog)$strain == "A" & colData(se_glog)$Next2 %in% c("B", "C"))
se_subset <- se_glog[, idx]
```

### 2. **Initialize Results Table:**

Prepare an empty `data.frame` to store the following for each metabolite:

-   Metabolite name

-   P-value from LRT

-   Chi-squared statistic

-   Degrees of freedom

-   Log-likelihood of nested model

-   Log-likelihood of complex model

```{r}
lrt_results <- data.frame(
  metabolite = rownames(se_subset),
  p_value = NA_real_, 
  chisq = NA_real_,
  df = NA_integer_,
  logLik_nested = NA_real_,
  logLik_complex = NA_real_,
  stringsAsFactors = FALSE
)
```

### **For Each Metabolite: Fit Two Models and Compare**

For each metabolite (row of `se_subset`):

-   Extract the intensity values across all selected samples.

-   Create a model frame including:

    -`day` (experimental time point)

    `-Next2` (group: B or C)

    Then fit two models:

<!-- -->

-   Nested Model: `intensity ~ day`

-   Complex Model: `intensity ~ Next2 + day`

    Perform a Likelihood Ratio Test (`lrtest`) to see if including `Next2` improves the model significantly.

-   `tryCatch` is used to handle metabolites that might fail to fit a model (e.g., due to too many missing values).

-   Chi-squared statistic tests if the more complex model fits significantly better.

```{r}
for (i in seq_len(nrow(se_subset))) {
  metab_id <- rownames(se_subset)[i]
  intensity <- assay(se_subset)[i, ]
  
  model_df <- data.frame(
    intensity = intensity,
    day = colData(se_subset)$day,
    Next2 = colData(se_subset)$Next2
  )
  
  tryCatch({
    nested <- glm(intensity ~ day, data = model_df)
    complex <- glm(intensity ~ Next2 + day, data = model_df)
    
    lrt <- lrtest(nested, complex)
    
    lrt_results$p_value[i] <- lrt$`Pr(>Chisq)`[2]
    lrt_results$chisq[i] <- lrt$Chisq[2]
    lrt_results$df[i] <- lrt$Df[2]
    lrt_results$logLik_nested[i] <- as.numeric(logLik(nested))
    lrt_results$logLik_complex[i] <- as.numeric(logLik(complex))
    
  }, error = function(e) {
    lrt_results$p_value[i] <- NA
  })
}

```

### 4. **Adjust P-values for Multiple Testing:**

Use False Discovery Rate (FDR) correction to adjust for multiple comparisons across metabolites.

```{r}
# Adjust p-values
lrt_results$adj_p_value <- p.adjust(lrt_results$p_value, method = "fdr")

```

### 5. **Summarize and View Results:**

-   Round numerical values for readability.

-   Sort by adjusted p-value to identify the most significant metabolites.

```{r}
# Round and arrange
lrt_table <- lrt_results %>%
  mutate(across(c(p_value, adj_p_value, chisq, logLik_nested, logLik_complex), round, digits = 8)) %>%
  arrange(adj_p_value)

# View top 10 as a summary table
print(head(lrt_table, 10))

```

#### Visualising output

Chisquared values vs pvalues

-   X-axis: The LRT Chi-squared statistic.

-   Y-axis: −log10(p-value) from the same test.

    This visualizes the strength of the test:

-   Higher Chi-squared = stronger statistical evidence against the null model.

-   Higher −log10(p-value) = higher statistical significance.

-   It gives a sanity check: strong chi-sq should correspond to low p-values (upper right corner).

    **Confirms the LRT workflow is robust and behaved as expected**

```{r}
ggplot(lrt_results, aes(x = chisq, y = -log10(p_value))) +
  geom_point(alpha = 0.6) +
  theme_minimal() +
  labs(title = "LRT Statistics per Metabolite",
       x = "Chi-squared Statistic",
       y = "-log10(p-value)")

```

### Preparing results for post analysis

-   Convert the `lrt_results` object fully into a `data.frame`.

-   Rename key columns

```{r}
colnames(lrt_results)
# Fully convert to base data.frame
lrt_subset_df <- as.data.frame(lrt_results) %>%
  dplyr::rename(
    Feature = metabolite,
    LRT_pvalue = p_value,
    adj_pval = adj_p_value
  )
```

-   Transpose the expression matrix (Rows = samples, Columns = features (metabolites))

-   Ensure column names match feature IDs.

-   Extract a clean metadata table for these samples.

```{r}
# Get expression matrix from se_subset
expr_cols <- t(assay(se_subset))  # transpose: samples as rows
colnames(expr_cols) <- rownames(se_subset)  # makes features = column names
# Metadata for those samples
df_subset_clean <- as.data.frame(colData(se_subset))



```

### **Check Feature Overlap Between LRT Results and Expression Matrix**

-   Sanity check: ensure features from LRT results are present in the expression matrix.

-   This prevents downstream errors when calculating fold-changes.

```{r}
# Check overlap
intersected_features <- intersect(lrt_subset_df$Feature, colnames(expr_cols))

# Sanity check
length(intersected_features)  # should be > 0

```

### **Calculate Fold Changes (Group C vs Group B)**

For each intersected feature:

-   Compute the **mean intensity** for samples where `Next2 == "C"`.

-   Compute the **mean intensity** for samples where `Next2 == "B"`.

-   Subtract the means to get a simple **fold change** (C − B).

```{r}

# Now safely calculate fold changes
fc_vals <- sapply(intersected_features, function(feat) {
  mean(expr_cols[df_subset_clean$Next2 == "C", feat]) - 
  mean(expr_cols[df_subset_clean$Next2 == "B", feat])
})

```

### **Verify Feature Matching**

Confirming the features (metabolite IDs) from the lrt match the features available in the expression matrix (`expr_cols`).

```{r}
# head(lrt_subset_df$Feature)
# head(rownames(se_subset))
# head(colnames(expr_cols))  # These should match lrt_subset_df$Feature

expr_cols <- t(assay(se_subset))  # samples = rows, features = columns
colnames(expr_cols) <- rownames(se_subset)  # features = original metabolite names
intersected_features <- intersect(lrt_subset_df$Feature, colnames(expr_cols))
length(intersected_features)

```

### **Prepare Clean Data Structures**

The expression matrix is correctly formatted as a `data.frame` (not a matrix).

Sample metadata (`Next2` groups) is ready for subsetting during fold-change calculation.

```{r}
expr_cols <- as.data.frame(t(assay(se_subset)))  # transpose + coerce to data frame
colnames(expr_cols) <- rownames(se_subset)       # features = column names

df_subset_clean <- as.data.frame(colData(se_subset))  # sample metadata
```

Calculate fold changes (c vs b)

```{r}
fc_vals <- sapply(intersected_features, function(feat) {
  mean(expr_cols[df_subset_clean$Next2 == "C", feat], na.rm = TRUE) -
  mean(expr_cols[df_subset_clean$Next2 == "B", feat], na.rm = TRUE)
})

```

visualising the results

### volcano

```{r}
volcano_data <- lrt_subset_df %>%
  filter(Feature %in% names(fc_vals)) %>%
  mutate(log2FC = fc_vals[Feature],
         negLog10P = -log10(LRT_pvalue))

ggplot(volcano_data, aes(x = log2FC, y = negLog10P)) +
  geom_point(alpha = 0.6) +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "red") +
  labs(title = "Volcano Plot: Strain A (Next2: B vs C)",
       x = "Log2 Fold Change (C vs B)",
       y = "-Log10 p-value") +
  theme_minimal()

```

### Volcano plot with EnhancedVolcano

```{r}
if (!requireNamespace("EnhancedVolcano", quietly = TRUE)) {
  install.packages("BiocManager")
  BiocManager::install("EnhancedVolcano")
}

colnames(lrt_subset_df)
str(lrt_subset_df)
library(EnhancedVolcano)

# Step 1: Ensure Feature names are in both datasets
stopifnot(all(lrt_subset_df$Feature %in% names(fc_vals)))

# Step 2: Add fold change and transformed p-values directly
lrt_subset_df$log2FC <- fc_vals[lrt_subset_df$Feature]
lrt_subset_df$negLog10P <- -log10(lrt_subset_df$LRT_pvalue)

# Make a copy
volcano_data <- lrt_subset_df

# Rename column manually
colnames(volcano_data)[colnames(volcano_data) == "LRT_pvalue"] <- "pvalue"

# Filter out NA p-values
volcano_data <- volcano_data[!is.na(volcano_data$pvalue), ]


str(volcano_data)
head(volcano_data$pvalue)


# Create volcano plot with EnhancedVolcano
EnhancedVolcano(volcano_data,
                lab = volcano_data$Feature,
                x = 'log2FC',
                y = 'pvalue',
                title = 'Strain A (Next2 C vs B)',
                subtitle = 'Volcano plot using EnhancedVolcano',
                xlab = 'Log2 Fold Change (C vs B)',
                ylab = '-Log10 P-value',
                pCutoff = 0.05,
                FCcutoff = 1.0,
                pointSize = 2.5,
                labSize = 3.5,
                colAlpha = 0.7,
                col = c('grey30', 'forestgreen', 'royalblue', 'red2'))

```

PCA

Select the top 100 most variable features

```{r}
library(FactoMineR)
library(factoextra)
# Top 100 most variable features
top_var_feats <- names(sort(apply(expr_cols, 2, var), decreasing = TRUE))[1:100]

```

Preparing PCA matrix and running PCA

```{r}
# Matrix for PCA (samples = rows)
pca_matrix <- as.matrix(expr_cols[, top_var_feats])
rownames(pca_matrix) <- rownames(df_subset_clean)

# Run PCA
pca_res <- PCA(pca_matrix, graph = FALSE)

```

Extract explained variance for PC1 and PC2 and prepare groupd labels

```{r}
# Extract explained variance for PC1 and PC2
explained_var_top <- round(pca_res$eig[1:2, 2], 1)

# Define group labels
group_labels <- factor(df_subset_clean$Next2)
names(group_labels) <- rownames(df_subset_clean)

```

Generate and Customize PCA Plot

-   Color samples by group (`Next2`) and add 95% confidence ellipses.

-   Customize axis labels with the variance explained by PC1 and PC2.

```{r}
# Generate PCA plot object
pca_plot_top <- fviz_pca_ind(pca_res,
                              label = "none",
                              habillage = group_labels,
                              addEllipses = TRUE,
                              ellipse.type = "confidence",
                              palette = c("#00AFBB", "#E7B800"),
                              title = "PCA (Top 100 Features): Strain A (Next2 B vs C)")

# Add custom axis labels with variance percentages
pca_plot_top +
  xlab(paste0("PC1 (", explained_var_top[1], "%)")) +
  ylab(paste0("PC2 (", explained_var_top[2], "%)"))

```

Using all features

```{r}
# Create matrix with all features (samples = rows, features = columns)
pca_matrix_all <- as.matrix(expr_cols)
rownames(pca_matrix_all) <- rownames(df_subset_clean)

# Run PCA
pca_res_all <- PCA(pca_matrix_all, graph = FALSE)

# Define group labels (e.g., treatment groups)
group_labels_all <- factor(df_subset_clean$Next2)
names(group_labels_all) <- rownames(df_subset_clean)

# Extract variance explained from PCA result
explained_var <- round(pca_res_all$eig[1:2, 2], 1)  # % variance for PC1 and PC2

# Generate base PCA plot
pca_plot_all <- fviz_pca_ind(pca_res_all,
                              label = "none",
                              habillage = group_labels_all,
                              addEllipses = TRUE,
                              ellipse.type = "confidence",
                              palette = c("#00AFBB", "#E7B800"),
                              title = "PCA (All Features): Strain A (Next2 B vs C)")

# Customize axis labels using ggplot2
pca_plot_all +
  xlab(paste0("PC1 (", explained_var[1], "%)")) +
  ylab(paste0("PC2 (", explained_var[2], "%)"))


```

Heatmap

Top 25

```{r}
library(pheatmap)
# ----- Heatmap ----- -------------------------------
# Select the top 25 features based on LRT significance (lowest LRT_pvalue)
# Make a base R version of the subset
top_features_df <- lrt_subset_df[!is.na(lrt_subset_df$LRT_pvalue), ]

# Order by p-value
top_features_df <- top_features_df[order(top_features_df$LRT_pvalue), ]

# Take the top 25 Feature values
top_features <- top_features_df$Feature[1:25]


# Extract heatmap data for these features
heat_data <- expr_cols[, top_features]

# Create an annotation data frame for the samples using metadata
annotation_df <- data.frame(
  Next2 = df_subset_clean$Next2,
  day = df_subset_clean$day
)
rownames(annotation_df) <- rownames(df_subset_clean)


# Define custom colors for 'Next2' and 'Day'
custom_colors <- list(
  Next2 = c(B = "red", C = "blue"),  # B = red, C = blue
  Day = c(
    "2" = "lightgrey",  # grey
    "3" = "darkgrey",  # dark grey
    "4" = "black",  # almost black
    "5" = "#654321",  # sienna (brown)
    "6" = "sienna"   # dark brown
  )
)


#library(pheatmap)

# Extract top features (already defined earlier)
heatmap_matrix <- t(expr_cols[, top_var_feats])  # Transpose to make features = rows

# Optional: scale features across samples (z-score)
heatmap_scaled <- t(scale(t(heatmap_matrix)))  # Scale by row (feature)

# Build updated column annotations
annotation_col <- data.frame(
  Next2 = df_subset_clean$Next2,
  Day = df_subset_clean$day,
  row.names = rownames(df_subset_clean)
)

# Plot heatmap with both annotations
pheatmap(heatmap_scaled,
         cluster_rows = TRUE,
         cluster_cols = TRUE,
         annotation_col = annotation_col,
         annotation_colors = custom_colors,
         show_rownames = FALSE,
         fontsize_col = 8,
         main = "Heatmap of Top Variable Features (Strain A, Next2 B vs C)")
```

All significant features heatmap

```{r}
# Select all features with adj p < 0.05
signif_features <- lrt_subset_df %>%
  filter(!is.na(adj_pval), adj_pval < 0.05) %>%
  pull(Feature)

# Extract and scale data
heatmap_matrix <- t(expr_cols[, signif_features])
heatmap_scaled <- t(scale(t(heatmap_matrix)))

# Reuse annotation_col from before
pheatmap(heatmap_scaled,
         cluster_rows = TRUE,
         cluster_cols = TRUE,
         annotation_col = annotation_col,
         annotation_colors = custom_colors,
         show_rownames = FALSE,
         fontsize_col = 8,
         main = "Heatmap of LRT-Significant Features (adj p < 0.05)")

```

All features

```{r}
# 1. Extract the full expression matrix (samples x features)
heatmap_matrix_all <- t(expr_cols)  # transpose: features as rows

# 2. Optional: scale each feature across samples (z-score)
heatmap_scaled_all <- t(scale(t(heatmap_matrix_all)))  # scale by row

# 3. Prepare annotations for samples (columns of the heatmap)
annotation_col <- data.frame(
  Next2 = df_subset_clean$Next2,
  Day = df_subset_clean$day,
  row.names = rownames(df_subset_clean)
)

# 4. Plot the full heatmap
pheatmap(heatmap_scaled_all,
         cluster_rows = TRUE,
         cluster_cols = TRUE,
         annotation_col = annotation_col,
         annotation_colors = custom_colors,
         show_rownames = FALSE,
         show_colnames = FALSE,
         fontsize_col = 8,
         main = "Heatmap of All Features (Strain A, Next2 B vs C)")

```

# IPA input for strain A next2 C vs B

Subset samples

```{r}
# Step 1: Subset to strain A, Next2 B or C
ipa_idx <- which(colData(se_glog)$strain == "A" & colData(se_glog)$Next2 %in% c("B", "C"))
se_ipa <- se_glog[, ipa_idx]

```

Extract expression data and metadata

```{r}
# Step 2: Extract expression matrix and sample info
expr <- assay(se_ipa)  # rows = features, cols = samples
metadata <- as.data.frame(colData(se_ipa))

```

Extract feature info

-   Feature-level metadata including `mzs` (mass-to-charge ratio) and `RTs` (retention time).

```{r}
# Step 3: Extract feature info — must contain mz and RT
feature_info <- as.data.frame(rowData(se_ipa))

# Safety check: must have mz and RT columns
if (!all(c("mzs", "RTs") %in% colnames(feature_info))) {
  stop("Feature metadata must include 'mz' and 'rt' columns.")
}

```

Building IPA export table

```{r}
# Step 4: Build the dataframe in ipaPy2 format
ipa_export <- data.frame(
  ids = as.numeric(rownames(expr)),  # or create sequential if rownames are not numeric
  mzs = feature_info$mzs,
  RTs = feature_info$RTs,
  expr,  # sample intensities
  check.names = FALSE
)
```

Standardising sample names and exporting to csv

```{r}
# Step 5: Reset column names for samples (e.g., sample1, sample2, ...)
sample_names <- paste0("sample", seq_len(ncol(expr)))
colnames(ipa_export)[4:ncol(ipa_export)] <- sample_names

# Step 6: Write to CSV
write.csv(ipa_export, "ipa_input_clusterFormat_strainA_C_vs_B.csv", row.names = FALSE)

```

# Rebuild the input for IPA (all metabolites, not just DE)

```{r}
# ✅ Final IPA Input Builder (Non-Clustered Format)

# Step 1: Subset relevant samples (Strain A, Next2 B and C)
ipa_idx_full <- which(colData(se_glog)$strain == "A" & colData(se_glog)$Next2 %in% c("B", "C"))
se_ipa_full <- se_glog[, ipa_idx_full]

# Step 2: Extract data
expr_all <- assay(se_ipa_full)
feature_info <- as.data.frame(rowData(se_ipa_full))

# Step 3: Ensure RTs are in seconds (if in minutes, convert)
if (mean(feature_info$RTs, na.rm = TRUE) < 100) {
  feature_info$RTs <- feature_info$RTs * 60
}

# Step 4: Create ipaPy2-ready dataframe
ipa_input <- data.frame(
  ids = rownames(expr_all),
  mzs = feature_info$mzs,
  RTs = feature_info$RTs,
  expr_all,
  check.names = FALSE
)

# Step 5: Rename intensity columns to sample1, sample2, ...
colnames(ipa_input)[4:ncol(ipa_input)] <- paste0("sample", seq_len(ncol(expr_all)))

# Step 6: Save for Python
write.csv(ipa_input, "ipa_input_allFeatures_pos_C_vs_B.csv", row.names = FALSE)

```
